---
title: "synthetic control code"
author: "Rebecca Johnson"
date: "February 1, 2017"
output: html_document
---

This code implements the placebo testing for synthetic control that is described in Abadie, Diamond, and Hainmmueller (2010): http://economics.mit.edu/files/11859

I describe each step in turn.

# Step one: iterate through each 'donor' unit and pretend it was the treatment unit

Donor units are observations (e.g., states; countries; schools) that are eligible to receive the treatment but that are not treated. The reason they're called donors is that when constructing the synthetic control, they form the pool of units that can be used for the composite/synthetic unit whose counterfactual outcomes we're comparing to the treatment state's outcomes. 

In this case, I am looking at policies that affect a state (WV in the below code) and so my donor units are states that were eligible to be treated by the policy WV treated to (only a subset of U.S. states since some had already switched to that treatment) but not treated. From now on, rather than referring to treatment and donor units, I'll refer to treatment and donor states.

The first step in placebo testing is to iterate through each of these donor states and pretend it was the treatment state and re-estimate whatever synthetic control model (e.g., same covariates; same pre-treatment v. post-treatment years) you ran on the actual treatment state. I do this using a loop.

```{r}
##load package
library(synth)

gapsplacebowv <- c() #empty vector to store gaps between tx state and control
treatpoolwv <- c(2:33) #indicators for all the donor states 

##loop iterates through the donor states
##and returns a data.frame with the gaps between the treatment
##and control for each donor state
for(i in treatpoolwv){
  allcontrols <- 2:33 #indices for all the donor states
  txindex <- which(allcontrols == i) #tells loop to treat state i as tx state for this iteration
  controlsnotx <- allcontrols[-txindex] #states not indexed below are the controls
  
  ##data preparation step for synth package
  ##note that the treatment.identifier is i and the controls are all
  ##indices other than i
  synthdataprepplacebo <- dataprep(foo = synthwithidwv,
                                   predictors = c("finalpercentspecialed", 
                                                  "stateunemployrate",
                                                  "singleparentpercent",
                                                  "leg_cont",
                                                  "totalpopinmil"),
                                   predictors.op = "mean",
                                   time.predictors.prior = 2000:2008,
                                   dependent = "finalpercentspecialed",
                                   unit.variable = "stateno",
                                   unit.names.variable = "statechar",
                                   time.variable = "year",
                                   treatment.identifier = i,
                                   controls.identifier = controlsnotx,
                                   time.optimize.ssr = 2000:2008,
                                   time.plot = 2000:2011)
  try(synthoutplacebo <- synth(data.prep.obj = synthdataprepplacebo)) 
  
  ##synthetic control outcome: weights each control states outcome by its weight 
  ycontrolplac <- synthdataprepplacebo$Y0plot %*% synthoutplacebo$solution.w
  
  ##treatment outcome
  ytreatedplac <- synthdataprepplacebo$Y1plot 
  
  ##for each state, creates a data.frame with the year in question
  #the outcome values for the synthetic control (ycontrolplace),
  ##the outcome values for the treatment (ytreatedplac), and 
  ##which donor state these outcomes refer to
  yearsplac <- seq(from = 2000, to = 2011, by = 1)
  statenames <- unique(synthwithidwv$statechar)
  stateplac <- rep(statenames[i], times = 12) 
  statedatawv <- cbind.data.frame(yearsplac, ycontrolplac, ytreatedplac, stateplac,
                                  stringsAsFactors = FALSE)
  colnames(statedatawv) <- c("year", "control", "treated", "state treated") 
  
  ##binds the above data.frame for all donor states that the loop
  ##iterates through
  gapsplacebowv <- rbind.data.frame(gapsplacebowv, statedatawv)
}

##calculates gaps outside loop as treatment outcome - control outcome
##but can also be calculated inside
gapsplacebowv$gap <- gapsplacebowv$treated - gapsplacebowv$control 
colnames(gapsplacebowv)[4] <- "statetreated" 


```


# Step two: calculate the RMSPE ratios during the post/pre-policy periods for these placebo states

We then calculate two quantities:

1. Pre-policy root mean squared prediction error (RMSPE): this looks at the gap between the treatment and synthetic control unit in the pre-policy period; the synthetic control is weighted to minimize this gap and you want it to be as small as possible. More formally, we square each gap so that the direction of the prediction error does not matter but its magnitude does, take the mean of those squared gaps, and then the square root of that quantity. 
2. Post-policy RMSPE: if the policy has an effect on your outcome (in this case, special education placement rates), then the treatment and control units should diverge and this should be large
3. Ratio of post-policy RMSPE to pre-policy RMSPE


The larger the ratio between 1 and 2, the more the treatment state and its counterfactual started to diverge in outcomes after the policy. You've already calculated this ratio for the treatment state during the main part of teh analysis. Now, you want to calculate it for each of the donor states that, in the placebo testing, you're pretending was treated. The reason is that for a causal interpretation, you want the treatment state's ratio to be as large as possible. But since there's no standard for what a large enough ratio is to 'count' as a causal effect, one way to contextualize whether the divergence in outcomes between the treatment unit and control unit is large is to compare this ratio for the actual treatment unit to a ratio for a donor state pretending that its a treatment state for the purposes of placebo testing. 


If the policy did have an effect in the treatment state, we would expect the state's post to pre ratios to be among the largest in these placebo tests, since we expect the treatment to cause a divergence in the outcomes (in this case, special education rates) for a treatment state in the post-policy period. 

The below function helps us calculate quantity 3 for each donor state.


```{r}
prepostmspe.function.mat.wv <- function(i) {
  state.names <- unique(gapsplacebowv$statetreated)
  state.group <- subset(gapsplacebowv, subset = (statetreated == state.names[i])) #subset to state i
  prepolstate <- state.group[state.group$year < 2009, ] #pre-policy outcomes
  rootpremspestate <- sqrt(mean(prepolstate$gap ^ 2)) #pre-policy rmspe
  postpolstate <- state.group[state.group$year > 2008, ]  #post-policy outcomes
  rootpostmspestate <- sqrt(mean(postpolstate$gap ^ 2)) #post-policy rmspe
  postpreratiostate <- rootpostmspestate / rootpremspestate #post/pre ratio
  mspestate <- cbind(rootpremspestate, rootpostmspestate, postpreratiostate) #bind above quantities
  return (mspestate)
}                

x <- seq(1:32) 
mspematwv <- sapply(x, prepostmspe.function.mat.wv) #apply function to all states
mspematdatawv <- as.data.frame(t(mspematwv)) 
rownames(mspematdatawv) <- unique(gapsplacebowv$statetreated) 
colnames(mspematdatawv) <- c("premspe", "postmspe", "postpreratio") 

```


The final step in placebo testing is to order the post-policy to pre-policy ratios from largest to smallest and calculate the treatment unit's rank in this ordered list. A lower rank means a larger gap post-policy gap relative to placebo units, and potentially a more causal interpretation of the policy. 
